Validation of Visual Statistical Inference, Applied to Linear Models

### 2012-06-28 paper_jasa.tex
model fitting details

Let $X_{\ell i} \sim Binom_{1, p_{\ell i}}$ be the random variable, with $X_{ \ell i}= 1$, if  subject $i$ correctly identifies the data plot from lineup $\ell$, and 0 otherwise.
We can model $p_{ \ell i}$ with a logistic regression with subject-specific random effects as:
\begin{equation} \label{mixed}
g( p_{\ell i} )= W_{\ell i} \delta +  Z_{\ell i}  \tau_{\ell i},  
\end{equation}

@@

\begin{equation}\label{power_exp} 
   Power(\beta)=Pr(p_{D} < p_0)  \quad \text{for}  \quad \beta \ne 0
\end{equation}
Figure \ref{fig:power_expected} shows the power of the conventional test in comparison to the expected power of the visual test obtained from Equation \ref{power_exp}. Notice that the expected power of the visual test is almost as good as the power of conventional test. \green{What do you mean by this? Where does this ``fact'' come from?} In fact, the only loss we experience stems from comparing the test statistic to only a discrete number of  $m-1$ samples from the null distribution.

%Theorem \ref{thm:power} helps to compute the expected power of visual inference.
%\begin{thm}\label{thm:power}
% Equation \ref{power_exp}  yields expected power of visual inference as Power($\beta$) = $E(1-p_D)^{m-1}$ .
%\end{thm}
% 
%{ \em proof}
%
%Since $p_0 \sim \text{Beta}(1,m-1)$,  for $t \in (0,1)$ we have the distribution function of $p_0$ be
%\begin{eqnarray*}
%F{p_0}(t) &=& (m-1) \int_0^t{(1-p_0)^{(m-2)}dp_0} \\
%  &=&  -(m-1) \int_1^{1-t}{x^{(m-2)}dx}\\
%  &=& \left[ x^{m-1}\right]^1_{1-t}\\
%  &=&1-(1-t)^{m-1} \rightarrow 1 \quad \text{as} \quad m \rightarrow \infty
%\end{eqnarray*}
%
%% verified this computation from http://www.wolframalpha.com using command int (m-1)*(1-p)^(m-2) dp, p=0,t
%
%Thus expected power in equation \ref{power_exp} be
%\begin{eqnarray*}
% Pr(p_{B} < p_0) &=& 1- Pr( p_0 \le p_{B} ) \\
%  &=& 1- \int_{0}^{1}{Pr( p_0 \le p_{B} |p_{B} =t) f_{p_{B} }(t)dt } \\
%  &=& 1-  \int_{0}^{1}{F{p_0}(t) f_{p_{B} }(t)dt } \\
%  &=& E(1-p_D)^{m-1}   \\
%%  & \rightarrow &1-  \int_{0}^{1}{ f_{p_{B} }(t)dt }  = F_{p_D}(0) \quad \text{as} \quad m \rightarrow \infty
%\end{eqnarray*}
%
%
\begin{figure}[hbtp]
   \centering
       \scalebox{.6}{\includegraphics{power_expected.pdf}}
       \caption{Expected power of visual test  and the power of conventional test for sample size $n =100$, $\sigma = 12$ and $m$=20. }
       \label{fig:power_expected}
\end{figure}

@@

\subsection{Model fitting}

The model in Equation \ref{mixed}is fit using \green{.... fill this in} Table \ref{tbl:model_par} shows the parameter estimates of the mixed effects model of the subject-specific variation. For all the three experiments, each subject saw a lineup just once. This doesn't allow fitting a lineup random effect, in addition to a subject random effect, but it does simplify the model since $\tau_{\ell
  i}=\tau_i$. The subject-specific probability ($p_i$) of successful evaluation of a lineup can be estimated in the model as $\hat p_i= ....$. So, we include a random interaction term with effect.

Different subjects are expected to have different skill levels in evaluating lineups. For this reason, when the subject-specific models (Equation \ref{mixed}) are fit to the data, in the next section, we set offset = $log(0.05/0.95)$ so that the estimated power has a fixed lower limit at 0.05 (Type-I error) when $E=0$. While fitting models (Equation \ref{mixed}) we set $W_{\ell i} = E_{\ell i}$, the effect of lineup $\ell$ that subject $i$ has evaluated. Also $E$ is used as the only fixed effect covariate without intercept but offset for experiment 1 and 2. For experiment 3 an intercept is estimated to allow the power curves to go above 0.05.

@@

\begin{table}[hbtp]
\caption{Parameter estimates of model \ref{mixed}.}
\begin{center}
\begin{tabular}{ccc}
  \hline
 &  \multicolumn{2}{c} {Estimates}  \\
 \cline{2-3}
 Experiment & Effect & inter subject correlation\\
  \hline
  1 & 0.39 &  \\ 
  2 & 1.22 &  \\ 
  3 &  & \\ 
   \hline
\end{tabular}
\end{center}
\label{tbl:model_par}
\end{table}

@@

For each parameter combination, the effect is computed as

\[
E=\beta*\sqrt {n}/\sigma.
\]

@@

%The power curves obtained by fitting model \ref{mixed} to the data for all the experiments applying different screening criteria is shown in figure \ref{fig:power_screening}. This shows how the screening criteria may affect the results. The corresponding power curves of conventional test are also shown. Notice that no mater what criteria we apply the result does not change much for experiment 2. Also, estimated power curve with criteria 5 for experiment 3 is above the conventional test power curve.

%
%\begin{figure}[hbtp]
%   \centering
%       \scalebox{0.70}{\includegraphics{power_mixed.pdf}}
%       \caption{The overall power estimated from model \ref{mixed} are shown. The corresponding power curve for conventional test is shown for comparison.}
%       \label{fig:power_mixed}
%\end{figure}

%Each participant was shown a sequence of 10 lineup plots.  In total, 3629 lineups were evaluated by 324 people coming from many different locations across the globe.  The results of the experiment are summarized in Figure \ref{fig:power_observed} which shows the observed power from the survey data calculated using Equation \ref{eqn:power_estimate} along with 95\% confidence interval calculated using Fisher's exact method.

@@

%\begin{figure*}[hbtp]
%   \centering
%       \scalebox{0.45}{\includegraphics{effect_power_subject_exp1.pdf}}
%       \scalebox{0.45}{\includegraphics{effect_power_subject_exp2.pdf}}
%       \caption{Subject specific power obtained from Equation \ref{eqn:effect} and the power of UMP test for different effect size $E$ as per experiment 1 and 2.}
%       \label{fig:power_effect_subject1}
%\end{figure*}

@@

%\begin{figure}[hbtp]
%   \centering
%       \scalebox{0.40}{\includegraphics{power_model.pdf}}
%       \caption{Estimated power curve from Equation \ref{eqn:power} along with 95\% confidence interval for sample size $n$ = 100 and $\sigma$ = 12.  The corresponding power curve for Uniformly Most powerful (UMP) test is shown for comparison.}
%       \label{fig:power_model}
%\end{figure}

%We fit model \eqref{mixed} to the survey data obtained from the simulation experiment. The estimated overall power curve obtained from equation \eqref{eqn:power} is shown in Figure \ref{fig:power_model}. Model \ref{mixed} also gives the subject specific power curves shown in Figure \ref{fig:power_subject}. The plot includes 20 randomly selected subject-specific power curves. Notice that the power curve estimated for one subject is above the UMP test power curve.

%\begin{figure}[hbtp]
%   \centering
%       \scalebox{0.45}{\includegraphics{power_subject.pdf}}
%       \caption{Estimated subject specific power curve from model \ref{mixed} for sample size $n$ = 100 and $\sigma$ = 12.  The corresponding power curve for Uniformly Most powerful (UMP) test is shown for comparison.}
%       \label{fig:power_subject}
%\end{figure}

@@

%Different subjects are expected to have different skill levels in evaluating lineups. This is accounted for in the mixed effects model (Equation \ref{mixed}).   
 The estimated subject-specific power curves are shown in Figure \ref{fig:power_mixed_subject} (light grey lines). The blue curve is the overall average power estimated from the model, and for comparison, the grey dashed lines are the power curves for the conventional test. Subject-specific power is quite different for three experiments. In experiment 1 subjects performed similarly, and substantially worse than the conventional test. In experiment 2 there is more variability between subjects, with some doing better than the conventional test on small effects. In experiment 3 there is the most subject-specific variation. Some subjects performed substantially better than the conventional test, and on average the visual test was better.

@@

\begin{figure}[hbtp]
   \centering
       \scalebox{0.60}{\includegraphics{power_mixed_subject.pdf}}
       \caption{Subject-specific  power estimated from model \ref{mixed} are shown. The corresponding power curve for conventional test (dashed line) is shown for comparison. The overall average power curve estimated from the model is shown (light blue).}
       \label{fig:power_mixed_subject}
\end{figure}

@@

With multiple observers it is possible to estimate the strength of the signal, using the proportion of correct picks. From Equation \ref{eqn:power_estimate}  we can write $1 - E[N_m]/K \approx (m-1)E[p_D] $. This gives the signal strength estimate as  
\[
\hat p_D= (1- {u}/{K})/({m-1})
\]
where $K$ is the total number of independent evaluations on a lineup and $u$ is the number of successful evaluations. Figure \ref{fig:pval_plot_signal} shows the plot signal strengths in relation with conventional test $p$-values, on a square root scale. For experiment 3, technically, there are no conventional $p$-values, but the $p$-values are computed in the same manner as experiment 2, assuming that the data satifies the assumptions. The blue line is a loess fit to the data, and the grey line indicates equality of the two axes. The relationship between the conventional $p$-value and the plot signal strength is very strong, particularly when the values are small. As the conventional $p$-value increases, the estimated signal strength increases -- observers find it harder to detect the actual data plot at larger $p$-values. The rise is very sharp for experiment 1, which indicates that observers have a harder time detecting the actual data plot for even small $p$-values.

@@

\begin{figure*}[hbtp]
   \centering
       \scalebox{0.6}{\includegraphics{p_val_plot_signal.pdf}}
       \caption{Conventional test $p$-value ($p_D$) vs plot signal strength ($\hat p_D$) estimated using Lemma \ref{lemma}. The solid line is a 45 degree line which represents the equity shown in equation \ref{eqn:power_estimate}. p-values are shown on square root scale scale. \blue{Is it better now?}}
    %   \hh{Mahbub, another change: could you change the y label to 'estimate of visual p-value $\hat{p_D}$ ?}
       \label{fig:pval_plot_signal}
\end{figure*}

@@

\begin{figure}[hbtp]
   \centering
       \scalebox{0.70}{\includegraphics{power_screening.pdf}}
       \caption{The overall power estimated from model \ref{mixed} are shown for various data sets obtained by applying 6 screening criteria.  The corresponding power curve for conventional test is shown in black solid lines for comparison.}
       \label{fig:power_screening}
\end{figure}

\begin{figure}[hbtp]
   \centering
       \scalebox{0.60}{\includegraphics{power_screening_subject.pdf}}
      \caption{The subject-specific power estimated from model \ref{mixed} are shown for various data sets obtained by applying 6 screening criteria.  The corresponding power curve for conventional test is shown in pink lines for comparison.}
       \label{fig:power_screening_subject}	
\end{figure}

>>>>

Let $X_{\ell i} \sim Binom_{1, p_{\ell i}}$ be the random variable, with $X_{ \ell i}= 1$, if  subject $i$ correctly identifies the data plot from lineup $\ell$, and 0 otherwise.
We can model $p_{ \ell i}$ with a logistic regression with subject-specific random effects as:
\begin{equation} \label{eqn:mixed}
g( p_{\ell i} )= W_{\ell i} \delta +  Z_{\ell i}  \tau_{\ell i},  
\end{equation}

@@

\begin{equation}\label{power_exp} 
   Power(\beta)=Pr(p_{D} < p_0)  \quad \text{for}  \quad \beta \ne 0
\end{equation}
Figure \ref{fig:power_expected} shows the power of the conventional test in comparison to the expected power of the visual test obtained from equation \eqref{power_exp}. Notice that the expected power of the visual test is almost as good as the power of conventional test. \green{What do you mean by this? Where does this ``fact'' come from?} In fact, the only loss we experience stems from comparing the test statistic to only a discrete number of  $m-1$ samples from the null distribution.

%Theorem \ref{thm:power} helps to compute the expected power of visual inference.
%\begin{thm}\label{thm:power}
% equation \eqref{power_exp}  yields expected power of visual inference as Power($\beta$) = $E(1-p_D)^{m-1}$ .
%\end{thm}
% 
%{ \em proof}
%
%Since $p_0 \sim \text{Beta}(1,m-1)$,  for $t \in (0,1)$ we have the distribution function of $p_0$ be
%\begin{eqnarray*}
%F{p_0}(t) &=& (m-1) \int_0^t{(1-p_0)^{(m-2)}dp_0} \\
%  &=&  -(m-1) \int_1^{1-t}{x^{(m-2)}dx}\\
%  &=& \left[ x^{m-1}\right]^1_{1-t}\\
%  &=&1-(1-t)^{m-1} \rightarrow 1 \quad \text{as} \quad m \rightarrow \infty
%\end{eqnarray*}
%
%% verified this computation from http://www.wolframalpha.com using command int (m-1)*(1-p)^(m-2) dp, p=0,t
%
%Thus expected power in equation \eqref{power_exp} be
%\begin{eqnarray*}
% Pr(p_{B} < p_0) &=& 1- Pr( p_0 \le p_{B} ) \\
%  &=& 1- \int_{0}^{1}{Pr( p_0 \le p_{B} |p_{B} =t) f_{p_{B} }(t)dt } \\
%  &=& 1-  \int_{0}^{1}{F{p_0}(t) f_{p_{B} }(t)dt } \\
%  &=& E(1-p_D)^{m-1}   \\
%%  & \rightarrow &1-  \int_{0}^{1}{ f_{p_{B} }(t)dt }  = F_{p_D}(0) \quad \text{as} \quad m \rightarrow \infty
%\end{eqnarray*}
%
%
\begin{figure}[hbtp]
   \centering
       \scalebox{.6}{\includegraphics{power_expected.pdf}}
       \caption{Expected power of visual test  and the power of conventional test for sample size $n =100$, $\sigma = 12$ and $m$=20. }
       \label{fig:power_expected}
\end{figure}

@@

\subsection{Model fitting}
For each parameter combination, the effect is computed as

\[
E=\beta (\sqrt {n}/\sigma).
\]

The model in equation \eqref{eqn:mixed} is fit using $E$ as the only fixed effect covariate without intercept. we set offset = $log(0.05/0.95)$ so that the estimated power has a fixed lower limit at 0.05 (Type-I error) when $E=0$. Different subjects are expected to have different skill levels in evaluating lineups. To account for subject specific variation we use subject as random effect. We also use an interaction between subject and effect $E$ to allow for subject specific random slope. This leads to setting  $W_{\ell i} = E_{\ell i}$, the effect of lineup $\ell$ that subject $i$ has evaluated in equation \eqref{eqn:mixed}.  While for experiment 1 and 2 we don't allow intercept but offset, for experiment 3 an intercept is estimated to allow the power curves to go above 0.05 when $E=0$. This is because forcing power to be fixed at 0.05 for $E=0$ is not justified by the experiment 3 data (figure \ref{fig:power_loess_effect}).

For all the three experiments, each subject saw a lineup just once. This doesn't allow fitting a lineup random effect, in addition to a subject random effect, but it does simplify the model since $\tau_{\ell i} = \tau_i$. The subject-specific probability ($p_i$) of successful evaluation of a lineup is then estimated using equation \eqref{eqn:mixed_power}. For computation we use R \citep{R} package lme4.

Table \ref{tbl:model_par} shows the parameter estimates of the mixed effects model of the subject-specific variation.

@@

\begin{table}[hbtp]
\caption{Parameter estimates of model \ref{eqn:mixed}.}
\begin{center}
\begin{tabular}{ccc}
  \hline
 &  \multicolumn{2}{c} {Estimates}  \\
 \cline{2-3}
 Experiment & Effect & inter subject correlation\\
  \hline
  1 & 0.39 &  \\ 
  2 & 1.22 &  \\ 
  3 &  & \\ 
   \hline
\end{tabular}
\end{center}
\label{tbl:model_par}
\end{table}

@@

@@

%The power curves obtained by fitting model \ref{eqn:mixed} to the data for all the experiments applying different screening criteria is shown in figure \ref{fig:power_screening}. This shows how the screening criteria may affect the results. The corresponding power curves of conventional test are also shown. Notice that no mater what criteria we apply the result does not change much for experiment 2. Also, estimated power curve with criteria 5 for experiment 3 is above the conventional test power curve.

%
%\begin{figure}[hbtp]
%   \centering
%       \scalebox{0.70}{\includegraphics{power_mixed.pdf}}
%       \caption{The overall power estimated from model \ref{eqn:mixed} are shown. The corresponding power curve for conventional test is shown for comparison.}
%       \label{fig:power_mixed}
%\end{figure}

%Each participant was shown a sequence of 10 lineup plots.  In total, 3629 lineups were evaluated by 324 people coming from many different locations across the globe.  The results of the experiment are summarized in Figure \ref{fig:power_observed} which shows the observed power from the survey data calculated using equation \eqref{eqn:power_estimate} along with 95\% confidence interval calculated using Fisher's exact method.

@@

%\begin{figure*}[hbtp]
%   \centering
%       \scalebox{0.45}{\includegraphics{effect_power_subject_exp1.pdf}}
%       \scalebox{0.45}{\includegraphics{effect_power_subject_exp2.pdf}}
%       \caption{Subject specific power obtained from equation \eqref{eqn:effect} and the power of UMP test for different effect size $E$ as per experiment 1 and 2.}
%       \label{fig:power_effect_subject1}
%\end{figure*}

@@

%\begin{figure}[hbtp]
%   \centering
%       \scalebox{0.40}{\includegraphics{power_model.pdf}}
%       \caption{Estimated power curve from equation \eqref{eqn:power} along with 95\% confidence interval for sample size $n$ = 100 and $\sigma$ = 12.  The corresponding power curve for Uniformly Most powerful (UMP) test is shown for comparison.}
%       \label{fig:power_model}
%\end{figure}

%We fit model \eqref{eqn:mixed} to the survey data obtained from the simulation experiment. The estimated overall power curve obtained from equation \eqref{eqn:power} is shown in Figure \ref{fig:power_model}. Model \ref{eqn:mixed} also gives the subject specific power curves shown in Figure \ref{fig:power_subject}. The plot includes 20 randomly selected subject-specific power curves. Notice that the power curve estimated for one subject is above the UMP test power curve.

%\begin{figure}[hbtp]
%   \centering
%       \scalebox{0.45}{\includegraphics{power_subject.pdf}}
%       \caption{Estimated subject specific power curve from model \ref{eqn:mixed} for sample size $n$ = 100 and $\sigma$ = 12.  The corresponding power curve for Uniformly Most powerful (UMP) test is shown for comparison.}
%       \label{fig:power_subject}
%\end{figure}

@@

%Different subjects are expected to have different skill levels in evaluating lineups. This is accounted for in the mixed effects model (Equation \ref{eqn:mixed}).   
 The estimated subject-specific power curves are shown in Figure \ref{fig:power_mixed_subject} (light grey lines). The blue curve is the overall average power estimated from the model, and for comparison, the grey dashed lines are the power curves for the conventional test. Subject-specific power is quite different for three experiments. In experiment 1 subjects performed similarly, and substantially worse than the conventional test. In experiment 2 there is more variability between subjects, with some doing better than the conventional test on small effects. In experiment 3 there is the most subject-specific variation. Some subjects performed substantially better than the conventional test, and on average the visual test was better.

@@

\begin{figure}[hbtp]
   \centering
       \scalebox{0.60}{\includegraphics{power_mixed_subject.pdf}}
       \caption{Subject-specific  power estimated from model \ref{eqn:mixed} are shown. The corresponding power curve for conventional test (dashed line) is shown for comparison. The overall average power curve estimated from the model is shown (light blue).}
       \label{fig:power_mixed_subject}
\end{figure}

@@

With multiple observers it is possible to estimate the strength of the signal, using the proportion of correct picks. From equation \eqref{eqn:power_estimate}  we can write $1 - E[N_m]/K \approx (m-1)E[p_D] $. This gives the signal strength estimate as  
\[
\hat p_D= (1- {u}/{K})/({m-1})
\]
where $K$ is the total number of independent evaluations on a lineup and $u$ is the number of successful evaluations. Figure \ref{fig:pval_plot_signal} shows the plot signal strengths in relation with conventional test $p$-values, on a square root scale. For experiment 3, technically, there are no conventional $p$-values, but the $p$-values are computed in the same manner as experiment 2, assuming that the data satifies the assumptions. The blue line is a loess fit to the data, and the grey line indicates equality of the two axes. The relationship between the conventional $p$-value and the plot signal strength is very strong, particularly when the values are small. As the conventional $p$-value increases, the estimated signal strength increases -- observers find it harder to detect the actual data plot at larger $p$-values. The rise is very sharp for experiment 1, which indicates that observers have a harder time detecting the actual data plot for even small $p$-values.

@@

\begin{figure*}[hbtp]
   \centering
       \scalebox{0.6}{\includegraphics{p_val_plot_signal.pdf}}
       \caption{Conventional test $p$-value ($p_D$) vs plot signal strength ($\hat p_D$) estimated using Lemma \ref{lemma}. The solid line is a 45 degree line which represents the equity shown in equation \eqref{eqn:power_estimate}. p-values are shown on square root scale scale. \blue{Is it better now?}}
    %   \hh{Mahbub, another change: could you change the y label to 'estimate of visual p-value $\hat{p_D}$ ?}
       \label{fig:pval_plot_signal}
\end{figure*}

@@

\begin{figure}[hbtp]
   \centering
       \scalebox{0.70}{\includegraphics{power_screening.pdf}}
       \caption{The overall power estimated from model \ref{eqn:mixed} are shown for various data sets obtained by applying 6 screening criteria.  The corresponding power curve for conventional test is shown in black solid lines for comparison.}
       \label{fig:power_screening}
\end{figure}

\begin{figure}[hbtp]
   \centering
       \scalebox{0.60}{\includegraphics{power_screening_subject.pdf}}
      \caption{The subject-specific power estimated from model \ref{eqn:mixed} are shown for various data sets obtained by applying 6 screening criteria.  The corresponding power curve for conventional test is shown in pink lines for comparison.}
       \label{fig:power_screening_subject}	
\end{figure}
